---
title: "4.3 — Pricing Strategies — Class Notes"
citeproc: false
bibliography: ../../static/bib/references.bib
csl: ../../static/bib/chicago-syllabus-no-bib.csl
output:
  blogdown::html_page:
    template: ../../pandoc/toc-title_html.template
    toc: true
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE,
                      message = FALSE,
                      warning = FALSE)
library(tidyverse)
```

*Monday, November 9, 2020*

# Overview

Today we continue our look at market power, by looking at the social consequences of market power: deadweight loss, allocative inefficiency, x-inefficiency, and rent-seeking; as well as the primary causes: barriers to entry of various sorts.

# Readings

See [today’s suggested readings](/reading/4.3-reading).

# Slides

- [<i class="fas fa-chalkboard-teacher"></i> View Lecture Slides](/slides/4.3-slides.html)
- [<i class="fas fa-file-pdf"></i> Download as PDF](/slides/4.3-slides.pdf)

# Practice Problems

Answers from last class’ [practice problems](/practice/4.1-practice).  are posted on that page.

# Assignments: Problem Set 5 (Due Sun Nov 8)

[Homework 5](/assignment/05-problem-set) (on 3.1-3.5) is due by PDF upload to Blackboard by 11:59 PM Sunday November 8. **This is likely your final graded homework**^[In the past, I have always released a HW 6 on Imperfect Competition just for practice for the Final Exam.]

# Appendix

## “I Know It When I See It” — How to Identify Market Power?


The only way to tell the difference between price discrimination and price differences due to costs in competitive markets (without actually being able to observe the firm’s marginal cost) is to find something that changes the price elasticity of demand without changing the cost. Price discrimination implies that a firm with market power sets its price based on the elasticity of demand and the marginal cost of produc- ing. Price in a competitive market depends only on marginal cost. (This is related to the distinction we discussed in Chapter 9 about how firms with market power react differently than competitive firms to rota- tions in demand.)

```{r}
demand_1=function(x){10-x}
mr_1=function(x){10-2*x}

demand_2=function(x){16-x}
mr_2=function(x){16-2*x}

mc_m=function(x){x+1}
mc_c=function(x){2*x+1}

pc<-ggplot(data=tibble(x=1:10), aes(x=x))+
  stat_function(fun=mc_m, geom="line", size=2, color="red")+
  geom_label(x=8,y=mc_m(8), color="red", label=expression(MC))+
  geom_hline(yintercept=7, size = 2, color = "blue")+
  geom_label(x=8,y=7, color="blue", label="Demand")+
  geom_segment(x=6,xend=6, y=0, yend=7, size=1, linetype="dotted")+
  scale_x_continuous(breaks = c(0,6),
                     labels = c(0,expression(q)),
                     limits = c(0,10),
                     expand=c(0,0))+
  scale_y_continuous(breaks = c(0,7),
                     labels = c(0,expression(p)),
                     limits = c(0,16),
                     expand=c(0,0))+
  theme_classic(base_family = "Fira Sans Condensed", base_size=12)+
  labs(x = "Quantity",
       y = "Price")

mp<-ggplot(data=tibble(x=1:10), aes(x=x))+
  stat_function(fun=demand_1, geom="line", size=2, color="blue")+
  stat_function(fun=mr_1, geom="line", size=2, color="purple")+
  stat_function(fun=mc_m, geom="line", size=2, color="red")+
  #stat_function(fun=mc_c, geom="line", size=2, color="red")+

  
  geom_label(x=8,y=demand_1(8), color="blue", label="Demand")+
  geom_label(x=4,y=mr_1(4), color="purple", label=expression(MR))+
  geom_label(x=8,y=mc_m(8), color="red", label=expression(MC))+
  #geom_label(x=4,y=mc_c(4), color="red", label=expression(MC[c]))+
  
  geom_segment(x=3,xend=3, y=0, yend=7, size=1, linetype="dotted")+
  geom_segment(x=0,xend=3, y=7, yend=7, size=1, linetype="dotted")+

  scale_x_continuous(breaks = c(0,3),
                     labels = c(0,expression(q)),
                     limits = c(0,10),
                     expand=c(0,0))+
  scale_y_continuous(breaks = c(0,7),
                     labels = c(0,expression(p)),
                     limits = c(0,14),
                     expand=c(0,0))+
  theme_classic(base_family = "Fira Sans Condensed", base_size=14)+
  labs(x = "Quantity",
       y = "Price")

library(patchwork)
pc+mp
```

### A Change in Marginal Cost

Perhaps as the result of a per-unit tax 

```{r}
mc_m2=function(x){x+1+2}
mc_c=function(x){2*x+1}

pc_c2<-pc+
  stat_function(fun=mc_m2, geom="line", size=2, color="darkred")+
  geom_label(x=8,y=mc_m2(8), color="darkred", label=expression(MC[2]))+
  geom_hline(yintercept=7, size = 2, color = "blue")+
  geom_label(x=8,y=7, color="blue", label="Demand")+
  geom_segment(x=4,xend=4, y=0, yend=7, size=1, linetype="dotted")+
  
  annotate("segment", x = 6, xend = 4, y = 2, yend = 2, colour = "black", size=2, alpha=1, arrow=arrow(length=unit(0.5,"cm"), ends="last", type="closed"))+

  scale_x_continuous(breaks = c(0,4,6),
                     labels = c(0,expression(q[2]),expression(q[1])),
                     limits = c(0,10),
                     expand=c(0,0))+
  scale_y_continuous(breaks = c(0,7),
                     labels = c(0,expression(p)),
                     limits = c(0,14),
                     expand=c(0,0))

mp_c2<-mp+
  stat_function(fun=mc_m2, geom="line", size=2, color="darkred")+
  geom_label(x=8,y=mc_m2(8), color="darkred", label=expression(MC[2]))+
  geom_segment(x=2.325, xend=2.325, y=0, yend=7.625, size=1, linetype="dotted")+
  geom_segment(x=0, xend=2.325, y=7.625, yend=7.625, size=1, linetype="dotted")+
  
  annotate("segment", x = 3, xend = 2.325, y = 2, yend = 2, colour = "black", size=2, alpha=1, arrow=arrow(length=unit(0.15,"cm"), ends="last", type="closed"))+
  annotate("segment", x = 0.5, xend = 0.5, y = 7, yend = 7.625, colour = "black", size=2, alpha=1, arrow=arrow(length=unit(0.25,"cm"), ends="last", type="closed"))+
  
  scale_x_continuous(breaks = c(0,2.325,3),
                     labels = c(0,expression(q[2]),expression(q[1])),
                     limits = c(0,10),
                     expand=c(0,0))+
  scale_y_continuous(breaks = c(0,7,7.625),
                     labels = c(0,expression(p[1]),expression(p[2])),
                     limits = c(0,14),
                     expand=c(0,0))
pc_c2+mp_c2
```

### Change in Demand

```{r}
demand_3=function(x){14-x}
mr_3=function(x){14-2*x}

pc_d2<-ggplot(data=tibble(x=1:10), aes(x=x))+
  stat_function(fun=mc_m, geom="line", size=2, color="red")+
  geom_label(x=8,y=mc_m(8), color="red", label=expression(MC))+
  geom_hline(yintercept=7, size = 2, color = "blue")+
  geom_label(x=3,y=7, color="blue", label="Demand")+
  geom_segment(x=6,xend=6, y=0, yend=7, size=1, linetype="dotted")+
  geom_hline(yintercept = 10, size = 2, color = "darkblue")+
  geom_label(x=3,y=10, color="darkblue", label=expression(Demand[2]))+
  geom_segment(x=9,xend=9, y=0, yend=10, size=1, linetype="dotted")+
  
  annotate("segment", x = 6, xend = 9, y = 2, yend = 2, colour = "black", size=2, alpha=1, arrow=arrow(length=unit(0.5,"cm"), ends="last", type="closed"))+
  annotate("segment", x = 1, xend = 1, y = 7, yend = 10, colour = "black", size=2, alpha=1, arrow=arrow(length=unit(0.5,"cm"), ends="last", type="closed"))+

    scale_x_continuous(breaks = c(0,6,9),
                     labels = c(0,expression(q[1]),expression(q[2])),
                     limits = c(0,10),
                     expand=c(0,0))+
  scale_y_continuous(breaks = c(0,7,10),
                     labels = c(0,expression(p[1]),expression(p[2])),
                     limits = c(0,16),
                     expand=c(0,0))+
  theme_classic(base_family = "Fira Sans Condensed", base_size=12)+
  labs(x = "Quantity",
       y = "Price")


mp_d2<-ggplot(data=tibble(x=1:10), aes(x=x))+
  stat_function(fun=demand_1, geom="line", size=2, color="blue", alpha = 0.35)+
  stat_function(fun=mr_1, geom="line", size=2, color="purple", alpha = 0.35)+
  stat_function(fun=mc_m, geom="line", size=2, color="red")+
  #stat_function(fun=mc_c, geom="line", size=2, color="red")+

  
  geom_label(x=8,y=demand_1(8), color="blue", label="Demand")+
  geom_label(x=4,y=mr_1(4), color="purple", label=expression(MR))+
  geom_label(x=8,y=mc_m(8), color="red", label=expression(MC))+
  #geom_label(x=4,y=mc_c(4), color="red", label=expression(MC[c]))+
  
  geom_segment(x=3,xend=3, y=0, yend=7, size=1, linetype="dotted")+
  geom_segment(x=0,xend=3, y=7, yend=7, size=1, linetype="dotted")+

  geom_segment(x=4.333,xend=4.333, y=0, yend=demand_3(4.333), size=1, linetype="dotted")+
  geom_segment(x=0,xend=4.333, y=demand_3(4.333), yend=demand_3(4.333), size=1, linetype="dotted")+

  stat_function(fun=demand_3, geom="line", size=2, color="darkblue")+
  geom_label(x=5,y=demand_3(5), color="darkblue", label=expression(Demand[2]))+
  stat_function(fun=mr_3, geom="line", size=2, color="purple")+
  geom_label(x=2.5,y=mr_3(2.5), color="purple", label=expression(MR[2]))+

  annotate("segment", x = 3, xend = 4.333, y = 3, yend = 3, colour = "black", size=2, alpha=1, arrow=arrow(length=unit(0.5,"cm"), ends="last", type="closed"))+
  annotate("segment", x = 0.5, xend = 0.5, y = 7, yend = demand_3(4.333), colour = "black", size=2, alpha=1, arrow=arrow(length=unit(0.5,"cm"), ends="last", type="closed"))+

  scale_x_continuous(breaks = c(0,3,4.333),
                     labels = c(0,expression(q[1]),expression(q[2])),
                     limits = c(0,10),
                     expand=c(0,0))+
  scale_y_continuous(breaks = c(0,7,demand_3(4.333)),
                     labels = c(0,expression(p[1]),expression(p[2])),
                     limits = c(0,14),
                     expand=c(0,0))+
  theme_classic(base_family = "Fira Sans Condensed", base_size=12)+
  labs(x = "Quantity",
       y = "Price")
pc_d2+mp_d2
```

## A Change in Price Elasticity of Market Demand

```{r}
demand_cr=function(x){7.75-0.5*x}

pc_e<-ggplot(data=tibble(x=1:10), aes(x=x))+
  stat_function(fun=demand_1, geom="line", size=2, color="blue", alpha =0.5)+
    geom_label(x=9,y=demand_1(9), color="blue", label="Demand")+
    stat_function(fun=mc_m, geom="line", size=2, color="red")+
    geom_label(x=9,y=mc_m(9), color="red", label="Supply")+
    stat_function(fun=demand_cr, geom="line", size=2, color="darkblue")+
    geom_label(x=9,y=demand_cr(9), color="darkblue", label=expression(Demand[E]))+

  geom_segment(x=4.5,xend=4.5, y=0, yend=5.5, size=1, linetype="dotted")+
  geom_segment(x=0,xend=4.5, y=5.5, yend=5.5, size=1, linetype="dotted")+
  scale_x_continuous(breaks = c(0,4.5),
                     labels = c(0,expression(q[c])),
                     limits = c(0,10),
                     expand=c(0,0))+
  scale_y_continuous(breaks = c(0,5.5),
                     labels = c(0,expression(p[c])),
                     limits = c(0,14),
                     expand=c(0,0))+
  theme_classic(base_family = "Fira Sans Condensed", base_size=12)+
  labs(x = "Quantity",
       y = "Price")

demand_r=function(x){8.5-0.5*x}
mr_r=function(x){8.5-1*x}

mp_e<-ggplot(data=tibble(x=1:10), aes(x=x))+
  stat_function(fun=demand_1, geom="line", size=2, color="blue",alpha=0.3)+
  stat_function(fun=mr_1, geom="line", size=2, color="purple",alpha=0.3)+
  stat_function(fun=mc_m, geom="line", size=2, color="red")+

  stat_function(fun=demand_r, geom="line", size=2, color="blue")+
  stat_function(fun=mr_r, geom="line", size=2, color="purple")+

  
  geom_label(x=8,y=demand_1(8), color="blue", label=expression(Demand[1]), alpha=0.3)+
  geom_label(x=4,y=mr_1(4), color="purple", label=expression(MR[1]), alpha=0.3)+
  geom_label(x=8,y=mc_m(8), color="red", label=expression(MC))+

  geom_label(x=8.5,y=demand_r(8.5), color="blue", label=expression(Demand[E]))+
  geom_label(x=4.25,y=mr_r(4.25), color="purple", label=expression(MR[E]))+

  
  geom_segment(x=3,xend=3, y=0, yend=7, size=1, linetype="dotted")+
  geom_segment(x=0,xend=3, y=7, yend=7, size=1, linetype="dotted")+

  geom_segment(x=3.75,xend=3.75, y=0, yend=6.5, size=1, linetype="dotted")+
  geom_segment(x=0,xend=3.75, y=6.5, yend=6.5, size=1, linetype="dotted")+

  annotate("segment", x = 3, xend = 3.75, y = 3, yend = 3, colour = "black", size=2, alpha=1, arrow=arrow(length=unit(0.2,"cm"), ends="last", type="closed"))+
  annotate("segment", x = 0.5, xend = 0.5, y = 6.5, yend = 7, colour = "black", size=2, alpha=1, arrow=arrow(length=unit(0.2,"cm"), ends="last", type="closed"))+

  scale_x_continuous(breaks = c(0,3,3.75),
                     labels = c(0,expression(q[c]), expression(q[m])),
                     limits = c(0,10),
                     expand=c(0,0))+
  scale_y_continuous(breaks = c(0,6.5,7),
                     labels = c(0,expression(p[m]), expression(p[c])),
                     limits = c(0,14),
                     expand=c(0,0))+
  theme_classic(base_family = "Fira Sans Condensed", base_size=12)+
  labs(x = "Quantity",
       y = "Price")

pc_e+mp_e
```

## Advanced Pricing Strategies

